Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-28 vertical-thermal-review.Rmd
update affiliations (Nidhi @ UCLA)
leaf traits paragraph
@NidhiVinod , I  put one line per sentence so its easier to see what I changed. But before I did that, I made two substantive edits. Please check. In particular, I'm not sure about the part between double asterisks. I thought this was driven by more stomatal closure.

Sun leaves tend to have higher photosynthetic rates per leaf area, made possible by higher nitrogen and chlorophyll concentrations per leaf area **and resulting in lower intercellular CO~2~ concentrations. **

In open canopies, where light is comparatively homogeneous, leaf

**Authors:** 
Nidhi Vinod^1^,
Martijn Slot^2^,
Ian McGregor^3^,
Elsa M. Ordway^4,5^,
Marielle N. Smith^6^,
Tyeen Taylor^7^,
Lawren Sack^5^,
Kristina J. Anderson-Teixeira^1,2^*

@@

2. Smithsonian Tropical Research Institute; Panama, Republic of Panama

3. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA

4. Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA 02138, USA

5. Department of Ecology and Evolutionary Biology, UCLA, Los Angeles, CA 90039, USA

@@

Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to increased reflectance, lower wettability, and/or higher $g_b$, resulting in lower $T_{leaf}$-$T_{air}$ and/or reduced water stress (Table `r table_leaf_traits`). Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed, with greater cuticle thickness, and higher trichome density [Fig. `r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a]. Sun leaves tend to have higher photosynthetic rates per leaf area, related to higher nitrogen and chlorophyll concentrations per leaf area and lower intercellular CO~2~ concentrations. Further, sun leaves tend to have higher stomatal and vein densities would enable their higher transpiration rates, and that would facilitate effective cooling  [@zwienieckiHydraulicLimitationsImposed2004]. Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], and, in addition to higher droplet contact angles, would have reduced wettability. Leaves in high light environments also exhibit greater leaf packing and clumping. In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance due to larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b]. In open canopies, where light is comparatively homogeneous, leaf traits may be more strongly related to maximum air temperature ($T_{air_{max}}$) and VPD stress than to light [@mediavillaFoliarPlasticityRelated2019a].

>>>>

**Authors:** 
Nidhi Vinod^1,2^,
Martijn Slot^3^,
Ian McGregor^4^,
Elsa M. Ordway^2,5^,
Marielle N. Smith^6^,
Tyeen Taylor^7^,
Lawren Sack^2^,
Kristina J. Anderson-Teixeira^1,3^*

@@

2. Department of Ecology and Evolutionary Biology, UCLA, Los Angeles, CA 90039, USA

3. Smithsonian Tropical Research Institute; Panama, Republic of Panama

4. Center for Geospatial Analytics; North Carolina State University; Raleigh, NC 27607, USA

5. Department of Organismic and Evolutionary Biology, Harvard University, Cambridge, MA 02138, USA

@@

Sun leaves have anatomical, morphological, and physiological traits that reduce heat due to increased reflectance, lower wettability, and/or higher $g_b$, resulting in lower $T_{leaf}$-$T_{air}$ and/or reduced water stress (Table `r table_leaf_traits`). 
Thus, sun leaves are generally smaller, thicker, with higher leaf mass per area, and are more deeply lobed, with greater cuticle thickness, and higher trichome density [Fig. `r fig_leaf_T`, @leighInfluenceLeafSize2017; @vogelSunLeavesShade1968a; @sackHowStrongIntracanopy2006;@zwienieckiHydraulicLimitationsImposed2004;@mathurPhotosyntheticEfficiencySun2018a].
Sun leaves tend to have higher photosynthetic rates per leaf area, made possible by higher nitrogen and chlorophyll concentrations per leaf area and resulting in lower intercellular CO~2~ concentrations. 
Further, sun leaves tend to have higher stomatal and vein densities would enable their higher transpiration rates, and that would facilitate effective cooling  [@zwienieckiHydraulicLimitationsImposed2004]. 
Steeper leaf angles reduce radiation loads and thereby decrease $T_{leaf}-T_{air}$ [@niinemetsAdjustmentFoliageStructure1998; @ballMaintenanceLeafTemperature1988], and, in addition to higher droplet contact angles, would have reduced wettability. Leaves in high light environments also exhibit greater leaf packing and clumping. 
In contrast, shade leaves have traits that maximize light capture in the diffuse-light environment with sunflecks (e.g., lower LMA and higher light absorptance efficiency per unit biomass), but their greater thermal capacitance due to larger leaf size and lower transpiration makes them more prone to overheating than sun leaves [@leighInfluenceLeafSize2017; @casasSunShadeLeaves2011; @schymanskiStomatalControlLeaf2013b]. 
In open canopies, where light is comparatively homogeneous, leaf traits may be more shabed by maximum air temperature ($T_{air_{max}}$) and VPD stress than to light [@mediavillaFoliarPlasticityRelated2019a].
